Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-05-03 PAPER.tex
edited jackknife text to reflect the interpretation of the latest null tests made with a subset of baselines

As shown in Figure \ref{fig:ps1_data}, our re-analysis of the PAPER-64 power spectrum shows biases above the predicted noise level, particularly at low-$k$ values. As discussed in Section \ref{sec:BiasTypes}, this bias is
most likely attributable to foreground leakage. The cause for biases at higher $k$-values is more difficult to pinpoint.

Here we perform three null tests on PAPER-64 data that aim to isolate systematics in the data and verify 
that they are not attributable to EoR. Similar to in Section \ref{sec:JackknifeOverview}, we take jackknives along different axes of the dataset to produce multiple power spectra. We then difference them (i.e. the null test) to tease out excess variances.

The three results are shown in Figure \ref{fig:null}. Each test displays the differenced power spectrum between two halves of a jackknife. We take jackknives along three different axes:

\begin{itemize}
%\item Original Data (black): This is identical to the unweighted, revised PAPER-64 power spectrum in Figure \ref{fig:ps1_data} (one baseline type only) with weighting matrix $\textbf{I}$. There are clear detections at low $k$'s.
\item Baselines: We split our dataset into two halves, where each contains half of the total baselines used in the 
analysis. No baselines are repeated between the two datasets.
\item Sidereal Hour: We split our dataset into two halves based on LST, namely the first half (LSTs 0.5-4.6 hours) and second half (LSTs 
4.6-8.6 hours).
\item Day: We split our dataset into even and odd Julian dates. We form power spectra for each separately, allowing the cross-multiplication of ``even" with ``even", for example, for this null test only. If the same sky signal is in both the ``even" and ``odd" datasets, we expect it to cancel out. %However, allowing these cross-multiplications can in practice introduce a low level noise bias... 
\end{itemize}

\begin{figure*}
	\centering
	\includegraphics[width=0.47\textwidth]{plots/null_bls.pdf}
	\includegraphics[width=0.47\textwidth]{plots/null_lsts.pdf}
	\includegraphics[width=0.47\textwidth]{plots/null_eo.pdf}
	\caption{Differenced power spectrum results (with $2\sigma$ errors) for three null tests, where a jackknife is taken along the baseline axis (top left), LST axis (top right), and even/odd Julian date axis (bottom). The results shown are unweighted (no signal loss), where the power spectrum values plotted are computed from the difference between two power spectra produced on either side of the jackknife axis. The gray shaded region in each plot is the estimated $2\sigma$ theoretical noise limit given the parameters of each test. We find that there are no significant systematics that differ between the tests, though there are low level biases at high-$k$ values in both the LST test and even/odd test. All three tests, whether biased or not, have an extra variance that is strongly $k$-dependent. Residual foregrounds dominate at $k$-values near the horizon ($k\sim\pm 0.1$\,$h$ Mpc$^{-1}$), which is not surprising given that foregrounds are brightest in this region post-delay filtering.}
	\label{fig:null}
\end{figure*}

@@

In investigating Figure \ref{fig:null}, we focus on two main aspects of each test --- whether the error bars are consistent with zero, and whether they are consistent with thermal noise. The former indicates a `passing' null test where there is no bias along the jackknife axis. The latter reveals whether there is additional variance present that lies above the thermal noise.

We find that the baselines test is fully consistent with noise, as all error bars pass through zero. Therefore, we conclude that the baselines making up the two jackknives either do not contain bias, or contain similar amounts of bias. We suspect it is the former, though more thorough jackknives along this axis are needed to make this conclusion. For the LST test, $90\%$ of all the error bars pass through zero, which is slightly lower than the expected $95\%$. This indicates a low level bias that is LST-dependent, and likely caused by residual foregrounds that vary in LST. In other words, the two jackknife halves differ by an amount greater than thermal noise and do not completely cancel out the sky signal when differencing. The even/odd test shows $81\%$ of all error bars being consistent with zero, again indicating the presence of a low level bias. The cause of this particular systematic is not obvious, and requires further investigation that we do not carry out in this paper. Most importantly, a clean detection of EoR would have passed all three tests, which is clearly not what we see.

Finally, we note that all three null tests exhibit error bars that are strongly $k$-dependent. The errors are largest just outside the horizon ($k\sim\pm 0.1$\,$h$ Mpc$^{-1}$) which is expected since foregrounds are brightest in this region post-delay filtering. We also note that all three tests, whether biased or unbiased, result in power spectrum errors that are slightly larger than the estimated thermal noise (gray band) at high-$k$. This indicates an additive noise component that is increasing our errors (while not affecting whether they pass through zero or not).

There are a couple possible explanations for this. Although we expect each cross-multiplication that is used in power spectrum estimation to have independent noise, there is still the possibility of a noise-foreground coupling term that can introduce power. This is because cross-multiplications produce four additive terms --- a signal-squared term (where signal includes both foregrounds and EoR), two cross-terms between the signal and noise, and one noise-only term. When differencing two power spectra, we expect the signal-only term to subtract out for a ``passing" null test, and we expect the noise-only terms to be consistent with thermal noise. The cross-terms, however, can introduce extra variance that would vary with foreground strength, therefore being consistent with the shape of our results.  One additional explanation, specifically for the slightly large error bars seen at high $k$'s, could simply be that our estimate of the thermal noise is too low, as all three tests seem to imply a larger, but consistent, estimate in this high-$k$ regime. Regardless of a potentially larger thermal noise prediction, however, the large error bars near the horizon limit still signify the presence of excess variance in our data.

In this section we have presented the first jackknife and null tests from the PAPER experiment. While simple, they outline a framework that can be used in the future. The 21\,cm community is beginning to recognize the importance of these types of tests (\citealt{pober_et_al2016b}) in characterizing power spectra at the EoR level, and it is clear that future results will require more substantial and thorough tests.

>>>>

As shown in Figure \ref{fig:ps1_data}, our illustrative PAPER-64 power spectrum shows biases above the predicted noise level, particularly at low-$k$ values. As discussed in Section \ref{sec:BiasTypes}, this bias is
most likely attributable to foreground leakage. %The cause for biases at higher $k$-values is more difficult to pinpoint.

Here we perform three null tests on PAPER-64 data that aim to isolate systematics in the data and verify 
that our biases are not attributable to EoR. Similar to in Section \ref{sec:JackknifeOverview}, we take jackknives along different axes of the dataset to produce multiple power spectra. We then difference them (i.e. the null test) to tease out excess variances.

The three results are shown in Figure \ref{fig:null}. Each test displays the differenced power spectrum between two halves of a jackknife, where the plotted points are the differenced power spectrum values, and the plotted errors are the errors of the two dataset halves added in quadrature. The expected thermal noise level (gray shaded regions) is the thermal noise of each dataset added in quadrature as well. Constructing the tests as such ensures that we are probing whether the variances of each dataset differ by an amount consistent with the thermal noise.

We take jackknives along three different axes:

\begin{itemize}
%\item Original Data (black): This is identical to the unweighted, revised PAPER-64 power spectrum in Figure \ref{fig:ps1_data} (one baseline type only) with weighting matrix $\textbf{I}$. There are clear detections at low $k$'s.
\item Baselines: We split our dataset into two halves, where each contains half of the total baselines used in the 
analysis. No baselines are repeated between the two datasets.
\item Sidereal Hour: We split our dataset into two halves based on LST, namely the first half (LSTs 0.5-4.5 hours) and second half (LSTs 
4.5-8.6 hours).
\item Day: We split our dataset into even and odd Julian dates. We form power spectra for each separately, allowing the cross-multiplication of ``even" with ``even", for example, for this null test only. If the same sky signal is in both the ``even" and ``odd" datasets, we expect it to cancel out. %However, allowing these cross-multiplications can in practice introduce a low level noise bias... 
\end{itemize}

\begin{figure*}
	\centering
	\includegraphics[width=0.47\textwidth]{plots/null_bls.pdf}
	\includegraphics[width=0.47\textwidth]{plots/null_lsts.pdf}
	\includegraphics[width=0.47\textwidth]{plots/null_eo.pdf}
	\caption{Differenced power spectrum results (with $2\sigma$ errors) for three null tests, where a jackknife is taken along the baseline axis (top left), LST axis (top right), and even/odd Julian date axis (bottom). The results shown are unweighted (no signal loss), where the power spectrum values plotted are computed from the difference between two power spectra produced on either side of the jackknife axis. The gray shaded region in each plot is the estimated $2\sigma$ theoretical noise limit given the parameters of each test. We find that there are no significant systematics for $k > \pm 0.2$\,$h$ Mpc$^{-1}$ for all three tests. However, we find that all tests exhibit an extra variance at $k$-values near the horizon ($k\sim\pm 0.1$\,$h$ Mpc$^{-1}$), likely due to foreground-noise coupling terms when foregrounds are brightest. Additionally, we find that the LST null test is not fully consistent with zero, implying a bias that is LST dependent and likely caused by varying foregrounds.}
	\label{fig:null}
\end{figure*}

@@

In investigating Figure \ref{fig:null}, we focus on three main possibilities --- whether the data points and error bars are consistent with thermal noise (``passing"), whether the error bars are consistent with zero but not consistent with thermal noise (``passing but has an additional variance"), or whether the error bars are not consistent with zero at all (``failing"). We examine each case in the context of our results below.

Firstly, all three null tests display data points that lie within the thermal noise gray band for $k > \pm 0.2$\,$h$ Mpc$^{-1}$. In addition, all three null tests show error bars consistent with the thermal noise level for those same $k$'s. This implies that the two jackknife halves do not differ by an amount greater than the thermal noise (i.e. the baselines making up the two jackknives either do not contain bias, or contain similar amounts of bias; we suspect it is the former though more thorough jackknives along this axis are needed to make this conclusion). We deem these as ``passing" null tests for the specific jackknives taken (again, dividing up the data in a different way along the same axes may not yield the same results, so more thorough testing is needed to be sure).

The second null test possibility (error bars consistent with zero but not with thermal noise) is displayed by the $k$-values just outside the horizon ($k\sim\pm 0.1$\,$h$ Mpc$^{-1}$) for all three tests. This indicates an additive noise component that is increasing our errors. More specifically, although we expect each cross-multiplication that is used in power spectrum estimation to have independent noise, there is still the possibility of a noise-foreground coupling term that can introduce power. This is because cross-multiplications produce four additive terms --- a signal-squared term (where signal includes both foregrounds and EoR), two cross-terms between the signal and noise, and one noise-only term. When differencing two power spectra (each with their own four terms), we expect the signal-only term to subtract out for a ``passing" null test, and we expect the noise-only terms to be consistent with thermal noise. The cross-terms, however, can introduce extra variance that would vary with foreground strength. It is therefore not surprising that this effect is largest at $k$-values just outside the horizon, where we expect foregrounds to be brightest post-delay filtering.

Lastly, a third null test result is an error bar not consistent with zero. This is the case for the LST null test at $k\sim-0.15$\,$h$ Mpc$^{-1}$ and $k\sim-0.2$\,$h$ Mpc$^{-1}$. In such a case, the two jackknife halves differ by an amount greater than the thermal noise (i.e. the data point is not in the thermal noise band), yet they are each constrained tightly by individual error bars that when combined, are not consistent with zero. This result implies that there exists a low level bias that is LST-dependent, and likely caused by residual foregrounds that vary in LST. Again, it is not surprising that this type of bias occurs near the horizon limit.

%Most importantly, a clean detection of EoR would have passed all three tests, which is clearly not what we see.

%We also note that all three tests, whether biased or unbiased, result in power spectrum errors that are slightly larger than the estimated thermal noise (gray band) at high-$k$. This indicates an additive noise component that is increasing our errors (while not affecting whether they pass through zero or not).

%There are a couple possible explanations for this. Although we expect each cross-multiplication that is used in power spectrum estimation to have independent noise, there is still the possibility of a noise-foreground coupling term that can introduce power. This is because cross-multiplications produce four additive terms --- a signal-squared term (where signal includes both foregrounds and EoR), two cross-terms between the signal and noise, and one noise-only term. When differencing two power spectra, we expect the signal-only term to subtract out for a ``passing" null test, and we expect the noise-only terms to be consistent with thermal noise. The cross-terms, however, can introduce extra variance that would vary with foreground strength, therefore being consistent with the shape of our results.  One additional explanation, specifically for the slightly large error bars seen at high $k$'s, could simply be that our estimate of the thermal noise is too low, as all three tests seem to imply a larger, but consistent, estimate in this high-$k$ regime. Regardless of a potentially larger thermal noise prediction, however, the large error bars near the horizon limit still signify the presence of excess variance in our data.

In this section we have presented the first jackknife and null tests from the PAPER experiment. Unsurprisingly, they imply that our measurements are biased by foregrounds, and not the EoR signal (a clean detection of EoR would have passed all three tests). While simple, these tests outline a framework that can be used by future measurements. The 21\,cm community is beginning to recognize the importance of these types of tests (\citealt{pober_et_al2016b}) in characterizing power spectra at the EoR level, and it is clear that future results will require more substantial and thorough investigations of this nature.
